Evolution of pre-copulatory and post-copulatory inbreeding behaviour and associated

### 2016-12-31 ms_pre_post.Rmd
Tweak the introduction

Inbreeding avoidance has been widely hypothesised to evolve through pre-copulatory and post-copulatory mechanisms as a consequence of inbreeding depression in offspring viability and female multiple mating (polyandry). However, no theoretical framework exists from which to derive hypothese concerning joint evolution of pre-copulatory and post-copulatory inbreeding avoidance. Evolution of post-copulatory inbreeding avoidance necessitates polyandry, potentially leading to complex evolutionary dynamics among inbreeding strategies and female mating frequency, each of which might be affected by direct negative costs. Hence, we use individual-based modelling to quantify evolution of interacting pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry across multiple cost combinations given severe inbreeding depression. We find that evolution of post-copulatory inbreeding avoidance increases selection for polyandry, that selection for costly inbreeding strategies was weak or negligible given a non-costly alternative strategy. Established pre-copulatory inbreeding avoidance generally precluded evolution of polyandry and therefore inbreeding avoidance, but established post-copulatory inbreeding aviodance did not preclude pre-copulatory inbreeding avoidance from evolving. We thereby provide a general framework for understanding evolution of pre-copulatory and/or post-copulatory inbreeding avoidance across systems, and conclude that evolution of inbreeding avoidance phenotypes and associated polyandry are predicted to interact and be strongly affected by direct costs.

@@

Inbreeding, defined as reproduction between relatives, often greatly reduces the fitness of resulting inbred offspring [termed 'inbreeding depression'; @Charlesworth1999; @Keller2002; @Charlesworth2009]. Such strong inbreeding depression is widely hypothesised to drive evolution of inbreeding avoidance, which can be enacted through multiple reproductive decisions [@Parker1979; @Parker2006; @Pusey1996; @Szulkin2012]. From the perspective of a focal breeding female, inbreeding avoidance might be achieved by avoiding mating with related males (i.e., pre-copulatory inbreeding avoidance), and by biasing fertilisation towards unrelated males after mating has occurred (i.e., post-copulatory inbreeding avoidance). Evolution of post-copulatory inbreeding avoidance requires that females are polyandrous, defined as mating with multiple males during a single reproductive bout. Further, such polyandry could itself evolve because it allows females to mate with additional unrelated males following an initial mating with a relative, thereby facilitating pre-copulatory inbreeding avoidance even without any post-copulatory female choice or otherwise biased fertilisation (i.e., given a 'fair raffle' among sperm provided by females' mates). Polyandry can thereby simultaneously allow females to mate with less closely related males and create the opportunity for further inbreeding avoidance enacted through active post-copulatory choice. Polyandry, which might not increase female reproductive success and therefore requires explanation [@Parker2013], has consequently been widely hypothesised to evolve as an adaptation to allow females to avoid inbreeding and thereby increase offspring fitness [@Zeh1997; @Jennions2000; @Tregenza2002]. However, there is surprisingly little theory or modelling that critically examines such widely-invoked hypotheses, and hence examines the conditions under which pre-copulatory or post-copulatory inbreeding avoidance, or both strategies, are predicted to evolve, or how evolution of such strategies of inbreeding avoidance might feed back to affect underlying evolution of polyandry.

Despite a lack of non-verbal theory and resulting predictions, numerous studies on diverse species have found evidence for female inbreeding avoidance. However, only a few such studies have determined whether inbreeding avoidance is enacted through pre-copulatory or post-copulatory mechanisms. Among these few studies, pre-copulatory inbreeding avoidance has been documented in weevils, birds, and butterflies [e.g., @Kuriwada2011; @Kingma2013; @Fischer2015], and post-copulatory inbreeding avoidance has been documented in crickets [e.g., @Simmons2006; @Bretman2009]. An even smaller number of studies comprise sufficient evidence for mechanisms of both pre-copulatory and post-copulatory inbreeding avoidance in two well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015]. A lack of both pre-copulatory and post-copulatory inbreeding has been found in fruit flies [e.g., @Ala-Honkola2011; @Tan2012], and might be reasonably inferred where no evidence of inbreeding avoidance exists [e.g., @Reid2014; @Reid2015]. To our knowledge, only @Liu2014 consider both pre-copulatory and post-copulatory inbreeding avoidance jointly, finding evidence of pre-copulatory, but not post-copulatory, inbreeding avoidance in cabbage beetles (*Colaphellus bowringi*). Taken together, all of these studies demonstrate that diverse combinations of pre-copulatory and post-copulatory inbreeding avoidance occur in nature. However, there is as yet no framework that predicts what combinations should be favoured by selection when polyandry and pre-copulatory and post-copulatory inbreeding avoidance can all potentially evolve. Consequently, there is no framework that allows the diversity of existing empirical results to be rationalised or interpreted, and no clear hypotheses that could be tested with these studies through comparative analyses or that could be applied to future empirical studies of individual systems.

A recent genetically-explicit individual-based model examined the conditions under which polyandry is predicted to evolve slely due to selection stemming from pre-copulatory inbreeding avoidance [@Duthie]. Evolution of polyandry occurred only under highly restricted conditions, requiring that direct costs were small, that very few males were available for a female's initial mate chioce but many were available for additional mate choice(s), or that polyandry was conditionally expressed when a focal female was related to her initial mate [@Duthie]. Without these conditions, increasing polyandry tended to increase rather than decrease the overall degree of inbreeding that females experienced, and hence reduced offspring fitness. This occurred because, once pre-copulatory inbreeding avoidance evolved, polyandrous females had already chosen available unrelated males as their initial mates. Their additional mates, chosen from the remaining male population, were therefore increasingly likely to include relatives. Evolution of polyandry purely to facilitate pre-copulatory inbreeding avoidance, and hence the resulting degree of inbreeding avoidance, was consequently restricted [@Duthie].

@@

However, if post-copulatory inbreeding avoidance could also evolve, polyandrous females could further bias paternity towards unrelated males within their set of chosen mates. Post-copulatory inbreeding avoidance might consequently reduce the cost of polyandry stemming from the accumulation of related mates, potentially facilitating evolution of polyandry to avoid inbreeding under broader conditions, and driving further consequent evolution of pre-copulatory or post-copulatory mate choice strategies. However, once both pre-copulatory and post-copulatory inbreeding avoidance, and resulting polyandry, can evolve, the long-term dynamics of these reproductive strategies become difficult to conceptualise and predict. Strong inbreeding depression might drive initial evolution of both pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry. However, if evolution of polyandry and post-copulatory inbreeding avoidance renders pre-copulatory inbreeding avoidance redundant, or vice versa, then only one inbreeding avoidance strategy might be maintained in the long-term.

The few previous models that considered the evolution of inbreeding avoidance through mate choice (as opposed to dispersal) have implicitly or explicitly considered the fate of a rare allele underlying pre-copulatory inbreeding avoidance in a population initially fixed for random mating [e.g., @Parker1979; @Parker2006; @Duthie; @Duthie2016a]. Such models are useful for isolating the invasion fitness of this single strategy. However, when both pre-copulatory and post-copulatory strategies can affect realised inbreeding, it cannot be assumed that both strategies will simultaneously invade a randomly mating population, nor that the invasion fitness of one strategy will be independent of the existence of invasion fitness of the other strategy. For example, if pre-adaptation or rapid selection results in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then new alleles underlying polyandry and post-copulatory inbreeding avoidance might be precluded from invading a population because the phenotypic effect of such alleles on the degree of inbreeding could be negligible. Conversely, fixation of alleles underlying polyandry and post-copulatory inbreeding avodiance might eliminate positive selection on alleles underlying pre-copulatory inbreeding avoidance. New theory, guided by modelling that evaluates invasion dynamics given appropriate sets of evolutionary constraints, is therefore needed.

Ultimately, the absolute and relative frequencies of alleles underlying pre-copulatory and post-copulatory inbreeding avoidance and polyandry will depend not only on the magnitudes of positive indirect selection steming from reduced inbreeding depression in females' offspring, but also on the magnitudes of direct negative selection on resulting phenotypes (i.e., the direct fitness costs of expressing each reproductive strategy). Polyandrous females can pay costs of mate searching or mating, for example, because these activities increase predation risk [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice and consequent reluctance to mate with relatives might pay costs of increased risk of complete mating failure or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay energetic costs associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs of pre-copulatory and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly strategy might become fixed over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy. However, such trajectories will also depend on the indirect benefits of each strategy, which might also differ and vary in relation to the evolution of the other strategy. Dynamic models are consequently required to understand and predict evolutionary outcomes.

@@

We address three key questions for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry given strong inbreeding depression. These questions include: (1) Does evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance facilitate evolution of polyandry that is otherwise severely constrained by increasing relatedness of additional mates? (2) How do costs associated with polyandry and pre-copulatory and post-copulatory inbreeding avoidance affect evolutionary outcomes and, in particular, the long-term persistence of reproductive strategies given cost assymetry? Finally, (3) how is selection for initially rare pre-copulatory or post-copulatory inbreeding avoidance affected if the other strategy of inbreeding avoidance is already fixed? By addressing these questions, we provide a framework for exploring the simultaneous evolution of multiple reproductive strategies through the use of individidual-based modelling, and we derive hypotheses amenable to empirical tests within and across taxa.

>>>>

Inbreeding avoidance has been widely hypothesised to evolve through pre-copulatory and post-copulatory mechanisms as a consequence of inbreeding depression in offspring viability and female multiple mating (polyandry). However, no theoretical framework exists from which to derive hypotheses concerning joint evolution of pre-copulatory and post-copulatory inbreeding avoidance. Evolution of post-copulatory inbreeding avoidance necessitates polyandry, potentially leading to complex evolutionary dynamics among inbreeding strategies and female mating frequency, each of which might be affected by direct negative costs. We use individual-based modelling to quantify evolution of interacting pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry across multiple cost combinations given severe inbreeding depression. We find that evolution of post-copulatory inbreeding avoidance increases selection for polyandry, that selection for costly inbreeding strategies was weak or negligible given a non-costly alternative strategy. Established pre-copulatory inbreeding avoidance generally precluded evolution of polyandry and therefore inbreeding avoidance, but established post-copulatory inbreeding avoidance did not preclude pre-copulatory inbreeding avoidance from evolving. We thereby provide a general framework for understanding joint evolution of pre-copulatory and/or post-copulatory inbreeding avoidance, and conclude that evolution of inbreeding avoidance phenotypes and associated polyandry are predicted to interact and be strongly affected by direct costs.

@@

Inbreeding, defined as reproduction between relatives, often greatly reduces the fitness of resulting inbred offspring [termed 'inbreeding depression'; @Charlesworth1999; @Keller2002; @Charlesworth2009]. Such strong inbreeding depression is widely hypothesised to drive evolution of inbreeding avoidance, which can be enacted through multiple reproductive decisions [@Parker1979; @Parker2006; @Pusey1996; @Szulkin2012]. From the perspective of a focal breeding female, inbreeding avoidance might be achieved by avoiding mating with related males (i.e., pre-copulatory inbreeding avoidance), and by biasing fertilisation towards unrelated males after mating has occurred (i.e., post-copulatory inbreeding avoidance). Evolution of post-copulatory inbreeding avoidance requires that females are polyandrous, defined as mating with multiple males during a single reproductive bout. Further, such polyandry might itself be under selection because given inbreeding depression it allows females to mate with additional unrelated males following an initial mating with a relative, thereby facilitating pre-copulatory inbreeding avoidance even without any post-copulatory female choice or otherwise biased fertilisation (i.e., given a 'fair raffle' among sperm provided by females' mates). Polyandry can thereby simultaneously allow females to mate with less closely related males and create the opportunity for further inbreeding avoidance enacted through active post-copulatory choice. Polyandry, which might not increase female reproductive success and therefore requires explanation [@Parker2013], has consequently been widely hypothesised to evolve as an adaptation to allow females to avoid inbreeding and thereby increase offspring fitness [@Zeh1997; @Jennions2000; @Tregenza2002]. However, there is surprisingly little theory or modelling that critically examines such widely-invoked hypotheses, and hence examines the conditions under which pre-copulatory or post-copulatory inbreeding avoidance, or both strategies, are predicted to evolve, or how evolution of such strategies of inbreeding avoidance might feed back to affect underlying evolution of polyandry.

Despite a paucity of theory and resulting predictions, numerous empirical studies on diverse species have found evidence for female inbreeding avoidance. However, only a few such studies have determined whether inbreeding avoidance is enacted through pre-copulatory or post-copulatory mechanisms. Among these few studies, pre-copulatory inbreeding avoidance has been documented in weevils, birds, and butterflies [e.g., @Kuriwada2011; @Kingma2013; @Fischer2015], and post-copulatory inbreeding avoidance has been documented in crickets [e.g., @Simmons2006; @Bretman2009]. An even smaller number of studies comprise sufficient evidence for mechanisms of both pre-copulatory and post-copulatory inbreeding avoidance in two well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015]. A lack of both pre-copulatory and post-copulatory inbreeding has been found in fruit flies [e.g., @Ala-Honkola2011; @Tan2012], and might be reasonably inferred where no evidence of inbreeding avoidance exists [e.g., @Reid2014; @Reid2015]. To our knowledge, only @Liu2014 consider both pre-copulatory and post-copulatory inbreeding avoidance jointly, finding evidence of pre-copulatory, but not post-copulatory, inbreeding avoidance in cabbage beetles (*Colaphellus bowringi*). Taken together, all of these studies demonstrate that diverse combinations of pre-copulatory and post-copulatory inbreeding avoidance occur in nature. However, there is as yet no theory that predicts what combinations should be favoured by selection when polyandry and pre-copulatory and post-copulatory inbreeding avoidance can all potentially evolve. Consequently, there is no theory that allows the diversity of existing empirical results to be rationalised or interpreted, and no clear hypotheses that could be tested with these studies through comparative analyses or that could be applied to future empirical studies of individual systems.

A recent genetically-explicit individual-based model examined the conditions under which polyandry is predicted to evolve solely due to selection stemming from pre-copulatory inbreeding avoidance [@Duthie]. Evolution of polyandry occurred only under highly restricted conditions, requiring that direct negative selection (i.e., 'costs') on polyandry were small, that very few males were available for a female's initial mate choice but many were available for additional mate choice(s), or that polyandry was conditionally expressed when a focal female was related to her initial mate [@Duthie]. Without these conditions, increasing polyandry tended to increase rather than decrease the overall degree of inbreeding that females experienced, and hence reduced offspring fitness. This occurred because, once pre-copulatory inbreeding avoidance evolved, polyandrous females had already chosen available unrelated males as their initial mates. Their additional mates, chosen from the remaining male population, were therefore increasingly likely to include relatives. Evolution of polyandry purely to facilitate pre-copulatory inbreeding avoidance, and hence the resulting degree of inbreeding avoidance, was consequently restricted [@Duthie].

@@

However, if post-copulatory inbreeding avoidance could also evolve, polyandrous females could further bias paternity towards unrelated males within their set of chosen mates. Post-copulatory inbreeding avoidance might consequently reduce the cost of polyandry stemming from the accumulation of related mates, potentially facilitating evolution of polyandry to avoid inbreeding under broader conditions, and driving further consequent evolution of pre-copulatory or post-copulatory mate choice strategies. However, once both pre-copulatory and post-copulatory inbreeding avoidance, and resulting polyandry, can all evolve, the long-term dynamics of these reproductive strategies become difficult to conceptualise and predict. Strong inbreeding depression might drive initial evolution of both pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry. However, if evolution of polyandry and post-copulatory inbreeding avoidance renders pre-copulatory inbreeding avoidance redundant, or vice versa, then only one inbreeding avoidance strategy might be maintained in the long-term.

The few previous models that considered the evolution of inbreeding avoidance through mate choice (as opposed to dispersal) have implicitly or explicitly considered the fate of a rare allele underlying pre-copulatory inbreeding avoidance in a population initially fixed for random mating [e.g., @Parker1979; @Parker2006; @Duthie; @Duthie2016a]. Such models are useful for isolating the invasion fitness of this single strategy. However, when both pre-copulatory and post-copulatory strategies can affect realised inbreeding, it cannot be assumed that both strategies will simultaneously invade a randomly mating population, nor that the invasion fitness of one strategy will be independent of the existence of invasion fitness of the other strategy. For example, if pre-adaptation or rapid selection results in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then new alleles underlying polyandry and post-copulatory inbreeding avoidance might be precluded from invading a population because the phenotypic effect of such alleles on the degree of inbreeding could be negligible. Conversely, fixation of alleles underlying polyandry and post-copulatory inbreeding avoidance might eliminate positive selection on alleles underlying pre-copulatory inbreeding avoidance. New theory, guided by modelling that evaluates invasion dynamics given appropriate sets of evolutionary constraints, is therefore needed.

Ultimately, the absolute and relative frequencies of alleles underlying pre-copulatory and post-copulatory inbreeding avoidance and polyandry will depend not only on the magnitudes of positive indirect selection stemming from reduced inbreeding depression in females' offspring, but also on the magnitudes of direct negative selection on resulting phenotypes (i.e., the direct fitness costs of expressing each reproductive strategy). Polyandrous females can pay costs of mate searching or mating, for example, because these activities increase predation risk [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice and consequent reluctance to mate with relatives might pay costs of increased risk of complete mating failure or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay energetic costs associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs of pre-copulatory and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly strategy might become fixed over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy. However, such trajectories will also depend on the indirect benefits of each strategy, which might also differ and vary in relation to the evolution of the other strategy. Dynamic models are consequently required to understand and predict evolutionary outcomes.

@@

We address three key questions for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry given strong inbreeding depression. These questions include: (1) Does evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance facilitate evolution of polyandry that is otherwise severely constrained from increasing relatedness of additional mates? (2) How do costs associated with polyandry and pre-copulatory and post-copulatory inbreeding avoidance affect evolutionary outcomes and, in particular, the long-term persistence of reproductive strategies given cost asymmetry? Finally, (3) how is selection for initially rare pre-copulatory or post-copulatory inbreeding avoidance affected if the other strategy of inbreeding avoidance is already fixed? By addressing these questions, we provide a framework for exploring the simultaneous evolution of multiple reproductive strategies through the use of individidual-based modelling, and we derive hypotheses amenable to empirical tests within and across taxa.
